Concurrent Programming in Harmony

### 2020-12-29 book/paper.tex
Latest

The most common use for locks is in building concurrent data structures.
By way of example, we will demonstrate how to build a concurrent queue.
The \texttt{queue} module will have the following API:
\begin{itemize}
\item $q = \mathtt{Queue}()$: allocate a new queue;
\item $\mathtt{put}(q, v)$: add $v$ to the tail of the queue;
\item $r = \mathtt{get}(q)$: returns $r = ()$ if $q$ is empty or $r = (v,)$ (a singleton tuple)
if $v$ was at the head of the queue.
\end{itemize}
See \autoref{fig:queuetest} for a simple test program that uses the queue.

@@

\index{ring buffer}%
A \emph{bounded buffer} (aka \emph{ring buffer}) is
a queue with the usual \texttt{put/get} interface,
but implemented using a circular buffer
\index{circular buffer}%
of a certain length and two pointers:
the \emph{tail} points where new items are enqueued and the \emph{head}
points where items are dequeued.
If the buffer is full, the enqueur must wait; if the buffer is empty, the
dequeuer must wait.
This problem is known as the ``Producer/Consumer Problem'' and was
proposed by Dijkstra~\cite{EWD329}.

@@

Split Binary Semaphores (SBS) is a general technique to implement conditional
waiting.  It was originally proposed by
Tony Hoare and popularized by Edsger Dijkstra~\cite{EWD703}.
SBS is an extension of a critical section that is protected by a single
binary semaphore.
If there are $n$ \emph{waiting conditions},
then SBS uses $n+1$ binary semaphores to protect the critical section.
An ordinary critical section has no waiting conditions.
Reader/writer locks have two waiting conditions:
\begin{enumerate}
\item readers waiting for a writer to release the lock;
\item writers waiting for another writer or all readers to release the lock.
\end{enumerate}
Similarly, a bounded buffer has two waiting conditions:
\begin{enumerate}
\item consumers waiting for the buffer to be non-empty;
\item producers waiting for an empty slot in the buffer.
\end{enumerate}
So each will require 3 binary semaphores if the SBS technique is applied.

@@

When leaving the critical section, the thread must open exactly one
of the gates, but it does not have to be the gate that it used to enter
the critical section.
In particular, when a thread leaves the critical section it should
check for each waiting gate if its waiting condition hold and if there are
threads trying to get through the gate.  If there is such a gate,
it must select one and open that gate.  If there is no such gate,
it must open the main gate.

@@

Note that the following invariants holds:
\begin{itemize}
\item At any time, at most one gate is open;
\item If some gate is open, then no thread is in the critical section;
\item If some thread is in the critical section, all gates are closed;
\item At any time, at most one thread is in the critical section.
\end{itemize}

The main gate is implemented by a binary semaphore, initialized in the
released state (signifying that the gate is open).
The waiting gates each consist of a pair: a counter that counts how many
threads are waiting behind the gate, and a binary semaphore initialized
in the acquired state (signifying that the gate is closed).

We will illustrate the technique using the reader/writer problem.
\autoref{fig:RWsplitsema} shows code.
The first step is to enumerate all waiting conditions.
In the case of the reader/writer
problem, there are two: a thread that wants to read may have to wait for a
writer to leave the critical section, while a thread that wants to write may
have to wait until all readers have left the critical section or until a writer has left.

The state of a reader/writer lock thus consists of the following:
\begin{itemize}
\item \textit{nreaders}: the number of readers in the critical section;
\item \textit{nwriters}: the number of writers in the critical section;
\item \textit{mutex}: the ``main gate'' binary semaphore;
\item \textit{r\_gate}: the ``waiting gate'' used by readers, consisting of a binary semaphore and the number of readers waiting to enter;
\item \textit{w\_gate}: the ``waiting gate'' used by writers, similar to the readers' gate.
\end{itemize}

Each of the
\texttt{read\_acquire}, \texttt{read\_release},
\texttt{write\_acquire}, and \texttt{write\_release} methods must maintain
this state.
Each of these methods first has to acquire the \textit{mutex}
(i.e., enter the main gate).
\texttt{read\_acquire} and \texttt{write\_acquire} must check to see
if it has to wait.
If so, it increments the count associated with its respective gate,
opens a gate (using \texttt{release\_one}, and then blocks until
its waiting gate opens up.

@@

Let us examine \texttt{read\_acquire} more carefully.
First the method acquires the \textit{mutex}.
Then, in the case that the thread, say $p$,
finds that there is a writer in the critical section
($\mathit{nwriters > 0}$), it increments the counter associated with
the readers' gate, leaves the critical section (\texttt{release\_one}),
and then tries to acquire the semaphore associated with the gate.
This will cause it to block until some thread opens that gate.

@@

\section*{Exercises}
\begin{problems}
% \item Assume that you have only binary semaphores.
% Implement counting semaphores using the split binary semaphores technique.
\item Several of the instantiations of \texttt{release\_one()} in
\autoref{fig:RWsplitsema} can be replaced by simply releasing \textit{mutex}.
Which ones?
\item Optimize your solutions to \autoref{ex:qcontains} to use reader/writer locks.
\item \label{ex:bbsbs} Implement a solution to the producer/consumer problem
using split binary semaphores.
\item \label{ex:boundlock} Using busy waiting, implement a ``bound lock'' that allows
up to \texttt{M} threads to acquire it at the same time.\footnote{A bound lock is a restricted version of a \emph{counting} semaphore.}
A bound lock
with \texttt{M == 1} is an ordinary lock.
You should define a constant \texttt{M} and two methods:
\texttt{acquire\_bound\_lock()}
and \texttt{release\_bound\_lock()}.
(Bound locks are useful for situations where too many threads working
at the same time might exhaust some resource such as a cache.)
\item Write a test program for your bound lock
that checks that no more than \texttt{M} threads can acquire the
bound lock.
\item Write a test program for bound locks
that checks that up to \texttt{M} threads
can acquire the bound lock at the same time.
\item Implement a \emph{bounded stack}.  Method \texttt{push}(\textit{item})
should block if the stack is full.  Method \texttt{pop()} should block
until the stack is non-empty and then return an item.
Also implement test programs for your solution.
\item \label{ex:gpu} Implement a thread-safe \emph{GPU allocator} by modifying
\autoref{fig:gpu}.
There are \texttt{N} GPUs identified by the numbers
1 through \texttt{N}.  Method \texttt{gpuAlloc()} returns the identifier
of an available GPU, blocking if there is currently no available GPU.
Method \texttt{gpuRelease}(\textit{gpu}) releases the given GPU.  It never needs
to block.
\item With reader/writer locks,
concurrency can be improved if a thread \emph{downgrades} its write lock
to a read lock when its done writing but not done reading.  Add
a \texttt{downgrade} method to the code in \autoref{fig:RWsplitsema}.
% How easy or hard would it be to do the same for the code in
% \autoref{fig:rw2lock}?
% Note, you cannot simply release the
% write lock and then obtain the read lock, because the thread needs
% to remain in the critical section.
\item \label{ex:onelane} Cornell's campus features some one-lane bridges.
On a one-lane bridge,
cars can only go in one direction at a time. Consider northbound
and southbound cars wanting to cross a one-lane bridge.
The bridge allows arbitrary many cars, as long as they're going in the
same direction.
Implement a lock that observes this requirement using SBS.
Write methods \texttt{OLBlock()} to create a new ``one lane bridge'' lock,
\texttt{nb\_enter()} that a car must invoke before going northbound on
the bridge and \texttt{nb\_leave()} that the car must invoke after leaving
the bridge.  Similarly write \texttt{sb\_enter()} and \texttt{sb\_leave()}
for southbound cars.
\item Extend the solution to \autoref{ex:onelane} by implementing the
requirement that at most $n$ cars are allowed on the bridge.  Add $n$
as an argument to \texttt{OLBlock}.

@@

A \emph{property}
\index{property}%
is a set of traces.
If a program has a certain property, that means that the traces that that
program allows are a subset of the traces in the property.
So far we have pursued two properties: \emph{mutual exclusion}
and \emph{progress}.  The former is an example of a
\emph{safety property}---it prevents something ``bad'' from
happening, like a reader and writer thread both entering the
critical section.  The \emph{progress} property is an example
of a \emph{liveness property}---guaranteeing that something good
eventually happens.
Informally (and inexactly), progress states that if no threads
are in the critical section, then some thread that wants to enter
can.

Progress is a weak form of liveness.  It says that \emph{some}
thread can enter, but it does not prevent a scenario such as
the following.  There are three threads repeatedly trying to
enter a critical section using a spinlock.  Two of
the threads successfully keep entering, alternating, but the third
thread never gets a turn.  This is an example of
\texttt{starvation}.  With a spinlock, this scenario could
even happen with two threads.  Initially both threads
try to acquire the spinlock.  One of the threads is
successful and enters.  After the thread leaves, it immediately
tries to re-enter.  This state is identical to the initial
state, and there is nothing that prevents the same thread
from acquiring the lock yet again.

It is worth noting that Peterson's Algorithm (\autoref{ch:peterson})
does not suffer from starvation, thanks to the \texttt{turn} variable
that alternates between 0 and 1 when two threads are contending for
the critical section.

@@

Implementing a reader/writer lock with Hoare monitors is not quite so
straightforward, unfortunately.  When a writer releases the lock, it
has to choose whether to signal a reader or another writer.  For that
it needs to know if there is a reader or writer waiting.  The simplest
solution would be to peek at the counters inside the respective condition
variable, but that breaks the abstraction.  The alternative is for the
reader/writer implementation to keep track of that state explicitly,
which complicates the code.  Also, it requires a deep understanding of
the SBS method to remember to place a call \texttt{signal} in the
\texttt{read\_acquire} method that releases additional readers that
may be waiting to acquire the lock.

@@

Like C, Java, and Python, Harmony does not have built-in language
support for monitors, but provides library support through the
Harmony \texttt{synch} module.
\autoref{fig:cv} shows the implementation of
condition variables in the \texttt{synch} module.
\texttt{Condition}() creates a new condition variable.
It is represented by a dictionary containing
a bag of name tags of threads waiting on the condition variable.
(The \texttt{synchS} library instead uses a list of contexts.)

@@

\begin{itemize}
\item \texttt{False < True};
\item integers are ordered in the natural way;
\item atoms are lexicographically ordered;
\item program counters are ordered by their integer values;
\item dictionaries are first converted into a list of ordered (key, value)
pairs.  Then two dictionaries are lexicographically ordered by this
representation;
\item a set is first converted into an ordered list, then lexicographically
ordered;
\item an address is a list of atoms.  \texttt{None} is the empty list of atoms.
Addresses are lexicographically ordered accordingly;
\item contexts (\autoref{app:context}) are ordered first by name tag, then by program counter, then by hash.
\end{itemize}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\textit{lv} = [lv =]... $e$ & \textit{lv} is an lvalue and $e$ is an expression\\
\hline
\textit{lv} \texttt{[op]=} $e$ & \texttt{op} is one of \texttt{+}, \texttt{-},
\texttt{*}, \texttt{/}, \texttt{//}, \texttt{\%},
\texttt{\string\&}, \texttt{|}, \texttt{\string\^},
\texttt{and}, and \texttt{or}\\
\hline
\texttt{pass} & do nothing\\
\hline
\texttt{del} \textit{lv} & delete\\
\hline
\texttt{assert $b$ [, $e$]} & $b$ is a boolean.  Optionally report value of expression $e$\\
\hline
\texttt{const a = $e$} & \texttt{a} is a bounded variable, $e$ is a constant expression\\
\hline
\texttt{def m $a$: S} & \texttt{m} is an identifier, $a$ a bounded variable, \texttt{S} a list of statements\\
\hline
\texttt{let $a$ = $e$ [let ...]: S} & $a$ is a bounded variable, $e$ is an expression, \texttt{S} a list of statements\\
\hline
\texttt{if $b$: S else: S} & $b$ is a boolean, \texttt{S} a list of statements\\
\hline
\texttt{while $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
\hline
\texttt{await $b$} & $b$ is a boolean\\
\hline
\texttt{for $a$ in $e$: S} & $a$ is a bounded variable, $e$ is a set,
                            \texttt{S} a list of statements\\
\hline
\texttt{atomic: S} & \texttt{S} a list of statements\\
\hline
\texttt{spawn m $e$ [, $t$]} & \texttt{m} is a method,
$e$ is an expression, $t$ is a tag (an expression) \\
\hline
\texttt{trap m $e$} & \texttt{m} is a method and $e$ is an expression \\
\hline
\texttt{go $c$ $e$} & $c$ is a context, $e$ is an expression \\
\hline
\texttt{import m, ...} & \texttt{m} identifies a module \\
\hline
\texttt{from m import ...} & \texttt{m} identifies a module \\
\hline
\end{tabular}

@@

{\small
\begin{tabular}{|l|l|}
\hline
Address & compute address from two components \\
\hline
Apply & pop $m$ and $i$ and apply $i$ to $m$, pushing a value \\
\hline
Assert, Assert2 & pop $b$ and check that it is \texttt{True}.  Assert2 also pops value to print \\
\hline
AtomicInc/Dec & increment/decrement the atomic counter of this context \\
\hline
Bag & pop $n$ and $n$ values and push a bag \\
\hline
Continue & no-op (but causes a context switch) \\
\hline
Choose & choose an element from the set on top of the stack \\
\hline
Cut & pop set or dict and push minimum and remainder \\
\hline
Del [$v$] & delete shared variable $v$ \\
\hline
DelVar [$v$] & delete thread variable $v$ \\
\hline
Dict & pop $n$ and $n$ key/value pairs and push a dictionary \\
\hline
Dup & duplicate the top element of the stack \\
\hline
Frame \texttt{m} $a$ & start method \texttt{m} with arguments $a$,
initializing variables.  \\
\hline
Go & pop context and value, push value on context's stack, and add to context bag \\
\hline
Jump $p$ & set program counter to $p$ \\
\hline
JumpCond $e$ $p$ & pop expression and, if equal to $e$, set program counter to $p$ \\
\hline
Load [$v$] & push the value of a shared variable onto the stack \\
\hline
LoadVar [$v$] & push the value of a thread variable onto the stack \\
\hline
Move $i$ & move stack element at offset $i$ to top of the stack \\
\hline
$n$-ary \textit{op} & apply $n$-ary operator \textit{op} to the top $n$ elements on the stack \\
\hline
Pop & pop a value of the stack and discard it \\
\hline
Push $c$ & push constant $c$ onto the stack \\
\hline
ReadonlyInc/Dec & increment/decrement the read-only counter of this context \\
\hline
Return & pop return address, push \texttt{result}, and restore program counter \\
\hline
Set & pop $n$ and $n$ elements and push a set of the elements \\
\hline
SetIntLevel & pop $e$, set interrupt level to $e$, and push old interrupt level \\
\hline
Spawn & pop tag, argument, and method and spawn a new context \\
\hline
Split & pop tuple and push its elements \\
\hline
Stop [$v$] & save context into shared variable $v$ and remove from context bag \\
\hline
Store [$v$] & pop a value from the stack and store it in a shared variable \\
\hline
StoreVar [$v$] & pop a value from the stack and store it in a thread variable \\
\hline
Trap & pop interrupt argument and method \\
\hline
\end{tabular}
}

>>>>

The most common use for locks is in building concurrent data structures.
By way of example, we will demonstrate how to build a concurrent queue.
The \texttt{queue} module will have the following API:
\begin{itemize}
\item $q = \mathtt{Queue}()$: allocate a new queue;
\item $\mathtt{put}(q, v)$: add $v$ to the tail of the queue;
\item $r = \mathtt{get}(q)$: returns $r = \mathtt{None}$
if $q$ is empty or $r = v$ if $v$ was at the head of the queue.
\end{itemize}
See \autoref{fig:queuetest} for a simple test program that uses the queue.

@@

\index{ring buffer}%
A \emph{bounded buffer} (aka \emph{ring buffer}) is
a queue with the usual \texttt{put/get} interface,
but implemented using a circular buffer
\index{circular buffer}%
of a certain length and two pointers:
the \emph{tail} points where new items are enqueued and the \emph{head}
points where items are dequeued.
If the buffer is full, the enqueuer must wait; if the buffer is empty, the
dequeuer must wait.
This problem is known as the ``Producer/Consumer Problem'' and was
proposed by Dijkstra~\cite{EWD329}.

@@

The Split Binary Semaphore (SBS) approach is a general technique for
implementing conditional waiting.  It was originally proposed by
Tony Hoare and popularized by Edsger Dijkstra~\cite{EWD703}.
SBS is an extension of a critical section that is protected by a single
binary semaphore.
If there are $n$ \emph{waiting conditions},
then SBS uses $n+1$ binary semaphores to protect the critical section.
An ordinary critical section has no waiting conditions.
For example, a bounded buffer has two waiting conditions:
\begin{enumerate}
\item consumers waiting for the buffer to be non-empty;
\item producers waiting for an empty slot in the buffer.
\end{enumerate}
So it will require 3 binary semaphores if the SBS technique is applied.

@@

When leaving the critical section, the thread must open exactly one
of the gates, but it does not have to be the gate that it used to enter
the critical section.
In particular, when a thread leaves the critical section it should
check for each waiting gate if its waiting condition holds and if there are
threads trying to get through the gate.  If there is such a gate,
it must select one and open that gate.  If there is no such gate,
it must open the main gate.

@@

Note that the following invariants hold:
\begin{itemize}
\item At any time, at most one gate is open;
\item If some gate is open, then no thread is in the critical section;
\item If some thread is in the critical section, all gates are closed;
\item At any time, at most one thread is in the critical section.
\end{itemize}

The main gate is implemented by a binary semaphore, initialized in the
released state (signifying that the gate is open).
The waiting gates each consist of a pair: a counter that counts how many
threads are waiting behind the gate and a binary semaphore initialized
in the acquired state (signifying that the gate is closed).

We will illustrate the technique using the reader/writer problem.
\autoref{fig:RWsplitsema} shows code.
The first step is to enumerate all waiting conditions.
In the case of the reader/writer
problem, there are two: a thread that wants to read may have to wait for a
writer to leave the critical section, while a thread that wants to write may
have to wait until all readers have left the critical section or until a writer has left.
%
The state of a reader/writer lock thus consists of the following:
\begin{itemize}
\item \textit{nreaders}: the number of readers in the critical section;
\item \textit{nwriters}: the number of writers in the critical section;
\item \textit{mutex}: the ``main gate'' binary semaphore;
\item \textit{r\_gate}: the ``waiting gate'' used by readers, consisting of a binary semaphore and the number of readers waiting to enter;
\item \textit{w\_gate}: the ``waiting gate'' used by writers, similar to the readers' gate.
\end{itemize}

Each of the
\texttt{read\_acquire}, \texttt{read\_release},
\texttt{write\_acquire}, and \texttt{write\_release} methods must maintain
this state.
First they have to acquire the \textit{mutex}
(i.e., enter the main gate).
After acquiring the \textit{mutex},
\texttt{read\_acquire} and \texttt{write\_acquire} each must check to see
if it has to wait.
If so, it increments the count associated with its respective gate,
opens a gate (using method \texttt{release\_one}), and then blocks until
its waiting gate opens up.

@@

Let us examine \texttt{read\_acquire} more carefully.
First the method acquires the \textit{mutex}.
Then, in the case that the thread
finds that there is a writer in the critical section
($\mathit{nwriters > 0}$), it increments the counter associated with
the readers' gate, leaves the critical section (\texttt{release\_one}),
and then tries to acquire the semaphore associated with the gate.
This will cause it to block until some thread opens that gate.

@@

\section*{Exercises}
\begin{problems}
% \item Assume that you have only binary semaphores.
% Implement counting semaphores using the split binary semaphores technique.
\item Several of the instantiations of \texttt{release\_one()} in
\autoref{fig:RWsplitsema} can be replaced by simply releasing \textit{mutex}.
Which ones?
\item Optimize your solutions to \autoref{ex:qcontains} to use reader/writer locks.
\item \label{ex:bbsbs} Implement a solution to the producer/consumer problem
using split binary semaphores.
\item \label{ex:boundlock} Using busy waiting, implement a ``bound lock'' that allows
up to \texttt{M} threads to acquire it at the same time.\footnote{A bound lock is a restricted version of a \emph{counting} semaphore.}
A bound lock
with \texttt{M == 1} is an ordinary lock.
You should define a constant \texttt{M} and two methods:
\texttt{acquire\_bound\_lock()}
and \texttt{release\_bound\_lock()}.
(Bound locks are useful for situations where too many threads working
at the same time might exhaust some resource such as a cache.)
\item Write a test program for your bound lock
that checks that no more than \texttt{M} threads can acquire the
bound lock.
\item Write a test program for bound locks
that checks that up to \texttt{M} threads
can acquire the bound lock at the same time.
% \item Implement a \emph{bounded stack}.  Method \texttt{push}(\textit{item})
% should block if the stack is full.  Method \texttt{pop()} should block
% until the stack is non-empty and then return an item.
% Also implement test programs for your solution.
\item \label{ex:gpu} Implement a thread-safe \emph{GPU allocator} by modifying
\autoref{fig:gpu}.
There are \texttt{N} GPUs identified by the numbers
1 through \texttt{N}.  Method \texttt{gpuAlloc()} returns the identifier
of an available GPU, blocking if there is currently no available GPU.
Method \texttt{gpuRelease}(\textit{gpu}) releases the given GPU.  It never needs
to block.
\item With reader/writer locks,
concurrency can be improved if a thread \emph{downgrades} its write lock
to a read lock when its done writing but not done reading.  Add
a \texttt{downgrade} method to the code in \autoref{fig:RWsplitsema}.
(Similarly, you may want to try to implement an \emph{upgrade} of a
read lock to a write lock.)
% How easy or hard would it be to do the same for the code in
% \autoref{fig:rw2lock}?
% Note, you cannot simply release the
% write lock and then obtain the read lock, because the thread needs
% to remain in the critical section.
\item \label{ex:onelane} Cornell's campus features some one-lane bridges.
On a one-lane bridge,
cars can only go in one direction at a time. Consider northbound
and southbound cars wanting to cross a one-lane bridge.
The bridge allows arbitrary many cars, as long as they're going in the
same direction.
Implement a lock that observes this requirement using SBS.
Write methods \texttt{OLBlock()} to create a new ``one lane bridge'' lock,
\texttt{nb\_enter()} that a car must invoke before going northbound on
the bridge and \texttt{nb\_leave()} that the car must invoke after leaving
the bridge.  Similarly write \texttt{sb\_enter()} and \texttt{sb\_leave()}
for southbound cars.
\item Extend the solution to \autoref{ex:onelane} by implementing the
requirement that at most $n$ cars are allowed on the bridge.  Add $n$
as an argument to \texttt{OLBlock}.

@@

A \emph{property}
\index{property}%
is a set of traces.
If a program has a certain property, that means that the traces that that
program allows are a subset of the traces in the property.
So far we have pursued two properties: \emph{mutual exclusion}
and \emph{progress}.  The former is an example of a
\emph{safety property}---it prevents something ``bad'' from
happening, like a reader and writer thread both acquiring a
reader/writer lock.
The \emph{progress} property is an example
of a \emph{liveness property}---guaranteeing that something good
eventually happens.
Informally (and inexactly), progress states that if no threads
are in the critical section, then some thread that wants to enter
can.

Progress is a weak form of liveness.  It says that \emph{some}
thread can enter, but it does not prevent a scenario such as
the following.  There are three threads repeatedly trying to
enter a critical section using a spinlock.  Two of
the threads successfully keep entering, alternating, but the third
thread never gets a turn.  This is an example of
\emph{starvation}.  With a spinlock, this scenario could
even happen with two threads.  Initially both threads
try to acquire the spinlock.  One of the threads is
successful and enters.  After the thread leaves, it immediately
tries to re-enter.  This state is identical to the initial
state, and there is nothing that prevents the same thread
from acquiring the lock yet again.
%
(It is worth noting that Peterson's Algorithm (\autoref{ch:peterson})
does not suffer from starvation, thanks to the \texttt{turn} variable
that alternates between 0 and 1 when two threads are contending for
the critical section.)

@@

Implementing a reader/writer lock with Hoare monitors is not quite so
straightforward, unfortunately.  When a writer releases the lock, it
has to choose whether to signal a reader or another writer.  For that
it needs to know if there is a reader or writer waiting.  The simplest
solution would be to peek at the counters inside the respective condition
variable, but that breaks the abstraction.  The alternative is for the
reader/writer implementation to keep track of that state explicitly,
which complicates the code.  Also, it requires a deep understanding of
the SBS method to remember to place a call to \texttt{signal} in the
\texttt{read\_acquire} method that releases additional readers that
may be waiting to acquire the lock.

@@

Like C, Java, and Python, Harmony does not have built-in language
support for monitors, but provides library support through the
Harmony \texttt{synch} module.
\autoref{fig:cv} shows the implementation of
condition variables in the \texttt{synch} module.
\texttt{Condition}() creates a new condition variable.
It is represented by a dictionary containing
a bag of contexts of threads waiting on the condition variable.
(The \texttt{synchS} library instead uses a list of contexts.)

@@

\begin{itemize}
\item \texttt{False < True};
\item integers are ordered in the natural way;
\item atoms are lexicographically ordered;
\item program counters are ordered by their integer values;
\item dictionaries are first converted into a list of ordered (key, value)
pairs.  Then two dictionaries are lexicographically ordered by this
representation;
\item a set is first converted into an ordered list, then lexicographically
ordered;
\item an address is a list of atoms.  \texttt{None} is the empty list of atoms.
Addresses are lexicographically ordered accordingly;
\item contexts (\autoref{app:context}) are ordered first by name, then by program counter, then by the remaining content.
\end{itemize}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\textit{lv} = [lv =]... $e$ & \textit{lv} is an lvalue and $e$ is an expression\\
\hline
\textit{lv} \texttt{[op]=} $e$ & \texttt{op} is one of \texttt{+}, \texttt{-},
\texttt{*}, \texttt{/}, \texttt{//}, \texttt{\%},
\texttt{\string\&}, \texttt{|}, \texttt{\string\^},
\texttt{and}, and \texttt{or}\\
\hline
\texttt{pass} & do nothing\\
\hline
\texttt{del} \textit{lv} & delete\\
\hline
\texttt{assert $b$ [, $e$]} & $b$ is a boolean.  Optionally report value of expression $e$\\
\hline
\texttt{const a = $e$} & \texttt{a} is a bounded variable, $e$ is a constant expression\\
\hline
\texttt{def m $a$: S} & \texttt{m} is an identifier, $a$ a bounded variable, \texttt{S} a list of statements\\
\hline
\texttt{let $a$ = $e$ [let ...]: S} & $a$ is a bounded variable, $e$ is an expression, \texttt{S} a list of statements\\
\hline
\texttt{if $b$: S else: S} & $b$ is a boolean, \texttt{S} a list of statements\\
\hline
\texttt{while $b$: S} & $b$ is a boolean, \texttt{S} a list of statements\\
\hline
\texttt{await $b$} & $b$ is a boolean\\
\hline
\texttt{for $a$ in $e$: S} & $a$ is a bounded variable, $e$ is a set,
                            \texttt{S} a list of statements\\
\hline
\texttt{atomic: S} & \texttt{S} a list of statements\\
\hline
\texttt{spawn m $e$ [, $t$]} & \texttt{m} is a method,
$e$ is an expression, $t$ is the initial thread-local state (an expression) \\
\hline
\texttt{trap m $e$} & \texttt{m} is a method and $e$ is an expression \\
\hline
\texttt{go $c$ $e$} & $c$ is a context, $e$ is an expression \\
\hline
\texttt{import m, ...} & \texttt{m} identifies a module \\
\hline
\texttt{from m import ...} & \texttt{m} identifies a module \\
\hline
\end{tabular}

@@

{\small
\begin{tabular}{|l|l|}
\hline
Address & compute address from two components \\
\hline
Apply & pop $m$ and $i$ and apply $i$ to $m$, pushing a value \\
\hline
Assert, Assert2 & pop $b$ and check that it is \texttt{True}.  Assert2 also pops value to print \\
\hline
AtomicInc/Dec & increment/decrement the atomic counter of this context \\
\hline
Bag & pop $n$ and $n$ values and push a bag \\
\hline
Continue & no-op (but causes a context switch) \\
\hline
Choose & choose an element from the set on top of the stack \\
\hline
Cut & pop set or dict and push minimum and remainder \\
\hline
Del [$v$] & delete shared variable $v$ \\
\hline
DelVar [$v$] & delete thread variable $v$ \\
\hline
Dict & pop $n$ and $n$ key/value pairs and push a dictionary \\
\hline
Dup & duplicate the top element of the stack \\
\hline
Frame \texttt{m} $a$ & start method \texttt{m} with arguments $a$,
initializing variables.  \\
\hline
Go & pop context and value, push value on context's stack, and add to context bag \\
\hline
Jump $p$ & set program counter to $p$ \\
\hline
JumpCond $e$ $p$ & pop expression and, if equal to $e$, set program counter to $p$ \\
\hline
Load [$v$] & push the value of a shared variable onto the stack \\
\hline
LoadVar [$v$] & push the value of a thread variable onto the stack \\
\hline
Move $i$ & move stack element at offset $i$ to top of the stack \\
\hline
$n$-ary \textit{op} & apply $n$-ary operator \textit{op} to the top $n$ elements on the stack \\
\hline
Pop & pop a value of the stack and discard it \\
\hline
Push $c$ & push constant $c$ onto the stack \\
\hline
ReadonlyInc/Dec & increment/decrement the read-only counter of this context \\
\hline
Return & pop return address, push \texttt{result}, and restore program counter \\
\hline
Set & pop $n$ and $n$ elements and push a set of the elements \\
\hline
SetIntLevel & pop $e$, set interrupt level to $e$, and push old interrupt level \\
\hline
Spawn & pop initial thread-local state, argument, and method and spawn a new context \\
\hline
Split & pop tuple and push its elements \\
\hline
Stop [$v$] & save context into shared variable $v$ and remove from context bag \\
\hline
Store [$v$] & pop a value from the stack and store it in a shared variable \\
\hline
StoreVar [$v$] & pop a value from the stack and store it in a thread variable \\
\hline
Trap & pop interrupt argument and method \\
\hline
\end{tabular}
}
